---
title: Expanded access to publicly subsidised lisdexamfetamine treatment for adults with attention-deficit/hyperactivity disorder, an interrupted time series analysis
output: word_document
---

We will be examining time series between **01 Feb 2016 and 31 Jan 2022**. Our intervention date (change date) is **1 February 2021** when the subsidy listing for lisdexamfetamine was expanded to allow people who were diagnosed with ADHD in adulthood subsidised access, prior to this adults who were diagnosed in adulthood were not eligible to access subsidised Lisdexamfetamine but may have accessed the medicine via the private market at a higher cost. 

This file uses PBS 10% sample data, the Australian Government Services Australia External Request Evaluation Committee granted data access (Approval Numbers: MI7542/RMS1941).Direct access to the data and analytical files to other individuals or authorities is not permitted without the express permission of the approving human research ethics committees and data custodians. The New South Wales (NSW) Population and Health Services Research Ethics Committee approved this study (Approval Number: 2013/11/494). 

Previous code, has cleaned the data and extracted all dispensing records for approved and subsidised ADHD medicines in Australia between 2016 and 2021 for people >18 years. We excluded dexamfetamine dispensing records which indicated use was for Narcolepsy. 
We included the following medicines: Guanfacine (atc_code: C02AC02), Dexamfetamine (N06BA02), Methylphenidate (N06BA04), Atomoxetine (N06BA09), Lisdexamfetamine (N06BA12). 
    
It also uses publicly available data from Australian Bureau of Statistics, data builder (see comments)

This file has two parts. 
First, create tables for the incidence (at least 18 month lookback period) and prevalence of ADHD medicine use among adults in Australia for 2021. Estimates were presented *per 1000 adults*. 

Second, to conduct an interrupted time series analysis quantifying changes in monthly dispensing following the lisdexamfetamine listing change on 1 February 2021. 
This part creates multiple pdfs to examine monthly dispensing rates as well as exploring ARIMA models, statistics and residuals. Finally, it produces a pdf with the final figure (monthly predicted and observed dispensing rates) and the effect estimates (with 95% confidence intervals) for the change in step, ramp, and trend effects. 

*this will cause blank pages in a word knit.

```{r load_libs_trends, eval=TRUE, include=FALSE}
# load libraries
library(dplyr)
# ARIMA libraries tidyvert
library(fable)
library(tsibble)
library(feasts)
library(magrittr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(sessioninfo)
library(DBI)
library(data.table)
library(purrr)
library(flextable)
library(tidyr)
library(knitr)
library(gridExtra)
library(readr)
library(ftExtra)
```

```{r read_data}
session_info()

options(scipen = 999)
#supress warnings caused by group_by and summarise in dpylr;
options(dplyr.summarise.inform = FALSE)

#These two files load PBS data and extract ADHD medicine dispensings to dataframe called ADHD;
#source("Lisdexamfetamine-load.R")
source("Lisdexamfetamine-clean.R")
```

Report basic characteristics, number of adults and dispensings (extrapolated to population from 10% sample)

```{r characteristics}
# number of dispensings
ADHD_char <- ADHD[supp_date %between% c(as.Date("2016-02-01"), as.Date("2022-01-31")) & age > 17, ]

num <- ADHD_char %>%
  summarise(cnt = sum(scripts)) %>%
  mutate(cnt = formatC(cnt * 10, format = "d", big.mark = ","))

cat("number of scripts dispensed", pull(num), sep = " ")

persons <- ADHD_char %>%
  group_by(pat_id) %>%
  arrange(pat_id) %>%
  distinct(pat_id, .keep_all = TRUE) %>%
  ungroup()

num <- persons %>%
  summarise(cnt = n()) %>%
  mutate(cnt = formatC(cnt * 10, format = "d", big.mark = ","))
cat("number of persons dispensed", pull(num), sep = " ")

persons %>%
  group_by(sex) %>%
  summarise(n = n()) %>%
  flextable() %>%
  set_caption("number of males and females")
```

ABS data were extracted on 30.01.22 from
<https://explore.data.abs.gov.au/vis?tm=quarterly%20population&pg=0&df[ds]=ABS_ABS_TOPICS&df[id]=ERP_Q&df[ag]=ABS&df[vs]=1.0.0&hc[Frequency]=Quarterly&pd=2015-Q1%2C2022-Q2&dq=1.2%2B1%2B3.85%2B86%2B87%2B88%2B89%2B90%2B91%2B92%2B100%2B99%2B98%2B97%2B96%2B95%2B94%2B93%2B84%2B83%2B82%2B81%2B80%2B79%2B78%2B77%2B76%2B75%2B74%2B73%2B72%2B71%2B70%2B69%2B68%2B67%2B66%2B65%2B64%2B63%2B62%2B61%2B60%2B59%2B58%2B57%2B56%2B55%2B49%2B48%2B47%2B46%2B45%2B44%2B43%2B42%2B41%2B40%2B39%2B38%2B37%2B36%2B35%2B34%2B33%2B32%2B31%2B30%2B29%2B28%2B27%2B26%2B25%2B24%2B23%2B22%2B21%2B20%2B54%2B53%2B52%2B51%2B50%2B19%2B18.AUS.Q&ly[cl]=TIME_PERIOD>
Filters that were applied include:
# Measures: "Estimated Resident Population"
# Sex: "Males"; "Females"; "Persons"
# Age:  (83 items) -> individual age from 18 (inclusive) to 99 and 100 and older group
# Region: "Australia"
# Frequency: "Quarterly"
# Time period: "Start: 2015-Q1"; "End: 2022-Q2"

These data were stored locally

```{r pop_data}
ABS_ERP <- readr::read_csv("../data/ABS_ERP_22_23.01.30.csv", show_col_types = FALSE)

# Convert ABS data into a cleaned-up tibble

ERP <- ABS_ERP %>%
  select(`SEX: Sex`, `AGE: Age`, `TIME_PERIOD: Time Period`, OBS_VALUE) %>%
  separate(`AGE: Age`, c("age"), sep = ":", extra = "drop", convert = "true") %>%
  separate(`SEX: Sex`, c("sex", "sex_chr"), sep = ":", convert = "true") %>%
  rename(date = `TIME_PERIOD: Time Period`) %>%
  rename(value = OBS_VALUE) %>%
  mutate(date = zoo::as.yearqtr(date, format = "%Y-Q%q"), age_group = agegrp(age), year = year(date)) %>%
  arrange(sex, date, age)

# Create half-year (Q2) population series by year, sex and age group

ERP_prev <- ERP %>%
  filter(grepl("Q2", date)) %>%
  group_by(year, sex, age_group) %>%
  summarise(value = sum(value)) %>%
  mutate(sex = case_when(
    sex == 1 ~ "Male",
    sex == 2 ~ "Female",
    sex == 3 ~ "persons"
  ))

# Create quarterly population series by quarter and sex (all ages)

ERP_time <- ERP %>%
  group_by(date, sex) %>%
  summarise(value = sum(value)) %>%
  mutate(sex = case_when(
    sex == 1 ~ "Male",
    sex == 2 ~ "Female",
    sex == 3 ~ "persons"
  ), date = zoo::as.Date(zoo::as.yearmon(date, format = "%Y-%m-%d")))

# Create monthly population series by month and sex (all ages) by linear interpolation

erp <- tsibble(ERP_time %>% mutate(sex = factor(sex)) %>%
  mutate(date = yearmonth(date)), key = sex, index = date) %>%
  full_join(tibble(
    date = rep(seq(yearmonth("2015-01-01"), yearmonth("2022-07-01"), by = 1), 3),
    sex = c(rep("Male", 91), rep("Female", 91), rep("persons", 91))
  ), by = c("date", "sex")) %>%
  fill_gaps()

erpx <- erp %>%
  split(list(.$sex)) %>%
  map_dfr(., ~ .x %>%
    model(lm = TSLM(value ~ trend(
      knots = yearmonth(c(
        "2016-01-01", "2016-04-01", "2016-07-01", "2016-10-01",
        "2017-01-01", "2017-04-01", "2017-07-01", "2017-10-01",
        "2018-01-01", "2018-04-01", "2018-07-01", "2018-10-01",
        "2019-01-01", "2019-04-01", "2019-07-01", "2019-10-01",
        "2020-01-01", "2020-04-01", "2020-07-01", "2020-10-01",
        "2021-01-01", "2021-04-01", "2021-07-01", "2021-10-01",
        "2022-01-01", "2022-04-01", "2022-07-01"
      ))
    ))) %>%
    interpolate(erp))
```

\newpage
## Part 1: Prevalence and incidence of ADHD medicines

```{r annual_prevalence}

# First dispensing per person per year;
ADHD_1 <- ADHD %>%
  filter(age > 17) %>%
  group_by(pat_id, year) %>%
  filter(row_number() == 1) %>%
  select(pat_id, year, age, sex, supp_date, atc_code, medicine, age_group_2) %>%
  rename(age_group = age_group_2)

# Prevalence for overall use by year;
ADHD_prev_overall <- ADHD_1 %>%
  group_by(year) %>%
  summarise(n = n()) %>%
  mutate(sex = "persons")
# Prevalence for use by year and sex;
ADHD_prev_sex <- ADHD_1 %>%
  group_by(year, sex) %>%
  summarise(n = n())

# Create half-year (Q2) population series by year and sex (all ages)

ERP_overall <- ERP_prev %>%
  group_by(year, sex) %>%
  summarise(value = sum(value))

prevalence <- rbind(ADHD_prev_overall, ADHD_prev_sex) %>%
  left_join(ERP_overall, by = c("year", "sex")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, sex) %>%
  filter(year == 2021 & sex == "persons") %>%
  mutate(age_group = "Overall")


# prevalent user based on age group and year
# Estimated resident populations - by age and sex and year
# estimated prevalence by age group, sex and year
ADHD_3 <- ADHD %>%
  filter(age > 17) %>%
  group_by(pat_id, year, age_group_2) %>%
  filter(row_number() == 1) %>%
  select(pat_id, year, age, sex, supp_date, atc_code, medicine, age_group_2) %>%
  rename(age_group = age_group_2)

ADHD_prev_age_sex <- ADHD_3 %>%
  group_by(age_group, sex, year) %>%
  summarise(n = n())

ADHD_prev_age <- ADHD_3 %>%
  group_by(age_group, year) %>%
  summarise(n = n()) %>%
  mutate(sex = "persons")


prevalence_age_sex <- rbind(ADHD_prev_age_sex, ADHD_prev_age) %>%
  left_join(ERP_prev, by = c("age_group", "sex", "year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, sex, age_group) %>%
  filter(year == 2021)

prevalence_age_sex$sex <- factor(prevalence_age_sex$sex, levels = c("persons", "Male", "Female"))

###############################################################################################################################################

# prevalent user based on medicine and year

ADHD_2 <- ADHD %>%
  filter(age > 17) %>%
  group_by(pat_id, year, medicine) %>%
  filter(row_number() == 1) %>%
  select(pat_id, year, age_group_2, sex, supp_date, atc_code, medicine, class) %>%
  rename(age_group = age_group_2) %>%
  mutate(class = case_when(
    atc_code == "C02AC02" ~ "Nonstimulant",
    atc_code == "N06BA02" ~ "Stimulant",
    atc_code == "N06BA04" ~ "Stimulant",
    atc_code == "N06BA09" ~ "Nonstimulant",
    atc_code == "N06BA12" ~ "Stimulant",
    TRUE ~ "Other"
  ))


# estimated prevalence by medicine and year

ADHD_prev_med <- ADHD_2 %>%
  group_by(year, class, medicine) %>%
  summarise(n = n())

prevalence_med <- ERP_overall %>%
  filter(sex == "persons") %>%
  left_join(ADHD_prev_med, by = c("year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, medicine) %>%
  filter(year == 2021) %>%
  mutate(age_group = "Medicine") %>%
  rename(sex = medicine)


#################################################################################
ADHD_prev_class <- ADHD_2 %>%
  group_by(year, class) %>%
  summarise(n = n())

prevalence_class <- ERP_overall %>%
  filter(sex == "persons") %>%
  left_join(ADHD_prev_class, by = c("year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, class) %>%
  filter(year == 2021) %>%
  mutate(age_group = "Medicine") %>%
  rename(sex = class)

ERP_2021 <- ERP_overall %>%
  filter(year == 2021 & sex == "persons") %>%
  mutate(age_group = "Population", sex = "Population") %>%
  rename(rate = value) %>%
  mutate(rate = formatC(rate, format = "d", big.mark = ","))


sexratio_18 <- prevalence_age_sex %>%
  filter(sex != "persons" & age_group == "18--24 years") %>%
  separate(rate, "prev", sep = " ", extra = "drop") %>%
  select(age_group, prev, sex) %>%
  mutate(prev = as.numeric(prev))

sexratio_25 <- prevalence_age_sex %>%
  filter(sex != "persons" & age_group == "25+ years") %>%
  separate(rate, "prev", sep = " ", extra = "drop") %>%
  select(age_group, prev, sex) %>%
  mutate(prev = as.numeric(prev))

sex_ratio <- data.frame(
  rate = c(
    round(pull(sexratio_18[2, 2]) / pull(sexratio_18[1, 2]), digits = 1),
    round(pull(sexratio_25[2, 2]) / pull(sexratio_25[1, 2]), digits = 1)
  ),
  sex = c("Male to female sex ratio", "Male to female sex ratio"),
  age_group = c("18--24 years", "25+ years"),
  year = c(2021, 2021)
)


prevalence_table <- rbind(prevalence, prevalence_age_sex, prevalence_med, prevalence_class, ERP_2021, sex_ratio) %>%
  select(age_group, sex, rate) %>%
  arrange(age_group, sex)


prevalence_table$age_group <- factor(prevalence_table$age_group, levels = c("Overall", "18--24 years", "25+ years", "Medicine"))
prevalence_table$sex <- factor(prevalence_table$sex, levels = c("persons", "Male", "Female", "Male to female sex ratio", "Stimulant", "Dexamfetamine", "Lisdexamfetamine", "Methylphenidate", "Nonstimulant", "Atomoxetine", "Guanfacine", "Population"))


```




```{r annual_incidence}

min_date <- as.Date("01-01-2015")

#incident dispensings, lookback period of at least 18 months;
ADHD_1 <- ADHD %>%
  group_by(pat_id) %>%
  arrange(pat_id, supp_date) %>%
  filter(row_number() == 1) %>%
  filter(supp_date > as.Date(min_date + months(17))) %>% # at least 18-month period look back
  filter(age > 17) %>%
  select(pat_id, year, age, sex, supp_date, atc_code, medicine, class, age_group_2) %>%
  rename(age_group = age_group_2) %>%
  mutate(class = case_when(
    atc_code == "C02AC02" ~ "Nonstimulant",
    atc_code == "N06BA02" ~ "Stimulant",
    atc_code == "N06BA04" ~ "Stimulant",
    atc_code == "N06BA09" ~ "Nonstimulant",
    atc_code == "N06BA12" ~ "Stimulant",
    TRUE ~ "Other"
  ))


# incidence for overall use by year;
ADHD_inc_overall <- ADHD_1 %>%
  group_by(year) %>%
  summarise(n = n()) %>%
  mutate(sex = "persons")

# incidence for use by year and sex;
ADHD_inc_sex <- ADHD_1 %>%
  group_by(year, sex) %>%
  summarise(n = n())


incidence <- rbind(ADHD_inc_overall, ADHD_inc_sex) %>%
  left_join(ERP_overall, by = c("year", "sex")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, sex) %>%
  filter(year == 2021 & sex == "persons") %>%
  mutate(age_group = "Overall")


ADHD_inc_age_sex <- ADHD_1 %>%
  group_by(age_group, sex, year) %>%
  summarise(n = n())

ADHD_inc_age <- ADHD_1 %>%
  group_by(age_group, year) %>%
  summarise(n = n()) %>%
  mutate(sex = "persons")


incidence_age_sex <- rbind(ADHD_inc_age_sex, ADHD_inc_age) %>%
  left_join(ERP_prev, by = c("age_group", "sex", "year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, sex, age_group) %>%
  filter(year == 2021)

incidence_age_sex$sex <- factor(incidence_age_sex$sex, levels = c("persons", "Male", "Female"))

###################################################################################################################################################

# estimated incidence by medicine and year
### create a new incidence based on incident medicine use:
ADHD_2 <- ADHD %>%
  group_by(pat_id, medicine) %>%
  arrange(pat_id, supp_date) %>%
  filter(row_number() == 1) %>%
  filter(supp_date > as.Date(min_date + months(17))) %>% # 18-month period look back
  filter(age > 17) %>%
  select(pat_id, year, age, sex, supp_date, atc_code, medicine, class, age_group_2) %>%
  rename(age_group = age_group_2) %>%
  mutate(class = case_when(
    atc_code == "C02AC02" ~ "Nonstimulant",
    atc_code == "N06BA02" ~ "Stimulant",
    atc_code == "N06BA04" ~ "Stimulant",
    atc_code == "N06BA09" ~ "Nonstimulant",
    atc_code == "N06BA12" ~ "Stimulant",
    TRUE ~ "Other"
  ))

# estimated incidence by medicine and year

ADHD_inc_med <- ADHD_2 %>%
  group_by(year, class, medicine) %>%
  summarise(n = n())

incidence_med <- ERP_overall %>%
  filter(sex == "persons") %>%
  left_join(ADHD_inc_med, by = c("year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, medicine) %>%
  filter(year == 2021) %>%
  mutate(age_group = "Medicine") %>%
  rename(sex = medicine)


#################################################################################
ADHD_inc_class <- ADHD_2 %>%
  group_by(year, class) %>%
  summarise(n = n())

incidence_class <- ERP_overall %>%
  filter(sex == "persons") %>%
  left_join(ADHD_inc_class, by = c("year")) %>%
  mutate(n = n * 10, prev = round((n / value) * 1000, digits = 1), n = paste0("(", formatC(n, format = "d", big.mark = ","), ")")) %>%
  unite(rate, c("prev", "n"), sep = " ") %>%
  select(year, rate, class) %>%
  filter(year == 2021) %>%
  mutate(age_group = "Medicine") %>%
  rename(sex = class)

ERP_2021 <- ERP_overall %>%
  filter(year == 2021 & sex == "persons") %>%
  mutate(age_group = "Population", sex = "Population") %>%
  rename(rate = value) %>%
  mutate(rate = formatC(rate, format = "d", big.mark = ","))

sexratio_18 <- incidence_age_sex %>%
  filter(sex != "persons" & age_group == "18--24 years") %>%
 separate(rate, "prev", sep = " ", extra = "drop") %>%
  select(age_group, prev, sex) %>%
  mutate(prev = as.numeric(prev))

sexratio_25 <- incidence_age_sex %>%
  filter(sex != "persons" & age_group == "25+ years") %>%
separate(rate, "prev", sep = " ", extra = "drop") %>%
  select(age_group, prev, sex) %>%
  mutate(prev = as.numeric(prev))

sex_ratio <- data.frame(
  rate = c(
    round(pull(sexratio_18[2, 2]) / pull(sexratio_18[1, 2]), digits = 1),
    round(pull(sexratio_25[2, 2]) / pull(sexratio_25[1, 2]), digits = 1)
  ),
  sex = c("Male to female sex ratio", "Male to female sex ratio"),
  age_group = c("18--24 years", "25+ years"),
  year = c(2021, 2021)
)


incidence_table <- rbind(incidence, incidence_age_sex, incidence_med, incidence_class, ERP_2021, sex_ratio) %>%
  select(age_group, sex, rate) %>%
  arrange(age_group, sex)


incidence_table$age_group <- factor(incidence_table$age_group, levels = c("Overall", "18--24 years", "25+ years", "Medicine"))
incidence_table$sex <- factor(incidence_table$sex, levels = c("persons", "Male", "Female", "Male to female sex ratio", "Stimulant", "Dexamfetamine", "Lisdexamfetamine", "Methylphenidate", "Nonstimulant", "Atomoxetine", "Guanfacine", "Population"))


incidence_table %>%
  rename(rate_inc = rate) %>%
  left_join(prevalence_table, by = c("age_group", "sex")) %>%
  arrange(age_group, sex) %>%
  group_by(age_group) %>%
  as_flextable(hide_grouplabel = TRUE) %>%
  bold(i = ~ !is.na(age_group), bold = TRUE, part = "body") %>%
  bold(i = ~ sex == "persons" | sex == "Population") %>%
  italic(i = ~ sex == "Stimulant" | sex == "Nonstimulant") %>% #
  font(fontname = "sans", part = "all") %>%
  bold(bold = TRUE, part = "header") %>%
  set_header_labels(sex = "", rate_inc = "Incidence (n)", rate = "Prevalence (n)") %>%
  add_footer_lines("ADHD, attention-deficit/hyperactivity disorder") %>%
  autofit() %>% 
  add_header_lines("Table 1. Incidence and prevalence of ADHD medicine use per 1000 adults in Australia in 2021, by age, sex and ADHD medicine")
  
  
```

\newpage

## Part 2: Interrupted time series analysis of ADHD medicines from February 2016 to January 2022.
Setting up Tsibbles for time series analysis.
    
```{r set_disp_ts}

# using tidyverts: Tsibbles:
# Index is a variable with inherent ordering from past to present.
# Key is a set of variables that define observational units over time.
# Each observation should be uniquely identified by index and key.
# Each observational unit should be measured at a common interval, if regularly spaced

# Convert data to time series object - dispensings by supply date: for each age group and medicine
# monthly aggregate dispensing by sex, age and medicine ()

all <- ADHD %>%
  filter(age_group == "Over 18") %>%
  mutate(medicine = class) %>%
  group_by(medicine, mos) %>%
  summarise(cnt = sum(scripts)) %>%
  mutate(sex = "persons", date = yearmonth(mos))

all_sex <- ADHD %>%
  filter(age_group == "Over 18") %>%
  mutate(medicine = class) %>% # updating plot individual stimulants and non-stimulants together
  # rename(date = mos) %>%
  group_by(sex, medicine, mos) %>%
  summarise(cnt = sum(scripts)) %>%
  mutate(date = yearmonth(mos))


all <- rbind(all, all_sex)

ts <- tsibble(all, key = c(sex, medicine), index = date) %>% 
  fill_gaps() %>% 
  filter_index("2016-02"~ "2022-01")

# define change date
change_date <- "2021-02-01"

#pre-intervention data - to map pre-intervention trend
ts2 <- ts %>% 
  filter_index("2016-02"~ change_date) %>% 
  fill_gaps() %>%
  inner_join(erpx, by = c("sex", "date")) %>%
  mutate(rate = (cnt * 10000) / value, cnt = as.numeric(cnt * 10))

ramp <- ts %>%
  ungroup() %>%
  filter(medicine == "Dexamfetamine" & sex == "Female") %>%
  filter_index(change_date ~.) %>% #tsibble filter function - all records from change date until end
  select(date) %>%
  mutate(ramp = 1:n())

tss <- merge(ts, ramp, by = "date", all.x = TRUE) %>%
  mutate(
    step = ifelse(as.Date(date) >= as.Date(change_date), 1, 0),
    ramp = ifelse(is.na(ramp), 0, ramp)
  ) %>%
  as_tsibble(key = c(sex, medicine), index = date)

## update rate per 1000 population
ts <- tss %>%
  inner_join(erpx, by = c("sex", "date")) %>%
  mutate(rate = (cnt * 10000) / value, cnt = as.numeric(cnt * 10))

# Remove data point for APRIL 2020, dispensing outlier due to covid 19 lockdowns;

ts %<>% filter(date != yearmonth("2020-04-01")) %>%
  fill_gaps()

nest_ts <- ts %>%
  select(date, medicine, sex, rate, cnt, ramp, step) %>%
  group_by(sex, medicine) %>%
  mutate(
    # log data variables
    log_rate = log(rate)
  ) %>%
  group_by(sex, medicine) %>%
  nest()
```


Exploring first whether data should be transformed and which differences are needed (are the data autocorrelated and seasonal)

```{r STL, eval = TRUE, include = FALSE}
## plot STL trend prior to intervention

plot_stl <- ts2 %>% model(STL = STL(log(rate)))


plot_all <- plot_stl %>% mutate(
  plot = pmap(list(STL, medicine, sex), ~ ..1 %>%
    components() %>%
    autoplot() + labs(title = paste("Log rate STL plots for", ..2, ..3, sep = " ")))
)


for (i in 1:12) {
  print(plot_all$plot[[i]])
}


```


```{r plot_raw_season, warning=FALSE, eval = TRUE, include = FALSE}

#create function for plotting seasonal data and autocorrelation;
plot_fun <- function(nest_data, y) {
  nest_data %>%
    mutate(plot = pmap(list(data, medicine, sex), ~ gg_tsdisplay(..1, y = .data[[y]], plot_type = "season") +
      labs(title = paste(y, "plots for", ..2, ..3, sep = " "))))
}

#create function for testing autocorrelation;

uniroot_fun <- function(nest_data, y) {
  nest_data %>%
    mutate(
      map_df(data, ~ .x %>% features(.data[[y]], unitroot_ndiffs)),
      map_df(data, ~ .x %>% features(.data[[y]], unitroot_nsdiffs))
    ) %>%
    select(-data)
}


plot_vars <- c("log_rate")

for (vars in plot_vars) {
  plot_all <- plot_fun(nest_data = nest_ts, y = vars)
  ft <- uniroot_fun(nest_ts, vars) %>%
    flextable() %>%
    ## create plot friendly font
    font(fontname = "sans", part = "all") %>%
    autofit() %>%
    add_header_lines(paste("Number of suggested seasonal and nonseasonal differences based on uniroot test for ", vars))


  # create pdf files
  pdf(file = paste(Sys.Date(), "_Raw plots (with season) for", vars, "by medicine and sex.pdf"))


  for (i in 1:12) {
    plot.new()
    print(plot_all[["plot"]][[i]])
  }
  # print flextable inside loop
  plot(ft, method = "grob")
  dev.off()
}

```


```{r markdown_plot_raw, warning = FALSE}

for (i in 1:12) {
    plot.new()
    print(plot_all[["plot"]][[i]])
}

ft
```


Model selection: Creating arma models and checking residuals

```{r model_data}

model_ts <- ts %>%
  model(
    seasonal_arma_log = ARIMA(log(rate) ~ pdq(d = 0) + PDQ(D = 1) + step + ramp + trend(), stepwise = FALSE),
  )

# seasonal lag formula - based on https://robjhyndman.com/hyndsight/ljung-box-test/

# For non-seasonal time series, use h=\min(10, T/5)h=min(10,T/5).
# For seasonal time series, use h=\min(2m, T/5)h=min(2m,T/5).
## number of dates -1 as excluded data from APR 2020


seasonal_lag <- min(12 * 2, (nrow(nest_ts$data[[1]])) / 5)
seasonal_lag2 <- min(12 * 2, (nrow(nest_ts$data[[1]]) - 12) / 5)

#check model AICcs;

model_ts %>%
  glance() %>%
  arrange(medicine, sex, AICc) %>%
  as_tibble() %>%
  as_grouped_data(groups = c("medicine", "sex")) %>%
  flextable() %>%
  bold(j = 7)


# we expect the residuals to fail this test when we have sufficient data

map_data <- nest_ts %>%
  select(medicine, sex) %>%
  as.data.frame()

# function to create a tibble containing information on one model for each stratification (sex and medicine)
filter_fun <- function(mable_data, model_var) {
  tibble(
    # including columns sex and medicine
    medicine = map_data[[1]],
    sex = map_data[[2]],
    # extracting individual model and placing one mable in each row of tibble (essential for plotting residuals ect)
    models = map2(map_data[[1]], map_data[[2]], ~ mable_data %>%
      filter(medicine == .x, sex == .y) %>%
      select(one_of(model_var, vars = model_var))),
    ## tidying the models
    tidy = map(models, tidy),
    # augment based on broom
    augment = map(models, augment),
    # caculate number of parameters to define degree of freedom
    para = map(tidy, nrow),
    # ljungbox test -need to unnest test when extracting these for use
    lbjung = map2_dfr(augment, para, ~ .x %>%
      features(.innov, ljung_box, lag = seasonal_lag, dof = .y)),
    test2 = map2_dfr(augment, para, ~ .x %>%
      features(.innov, ljung_box, lag = seasonal_lag2, dof = .y)),
    # extract model specs
   model_name = map_chr(models, ~ .x[[1]][[1]] %>% model_sum() )
  )
}

```

```{r residual_plots, eval = TRUE, warning = FALSE, include = FALSE}

#plot residuals for models and table of ljung pvalue test 
#gg_tsresidual may only work on one single mable.

model <- c("seasonal_arma_log")


#for (model in models) {
  # plot function
  p <- filter_fun(model_ts, model)

  # create flextable object for residual test table
  ft <- p %>%
    select(medicine, sex, model_name, contains("test2")) %>%
    unnest(test2) %>%
    arrange(sex) %>%
    as_grouped_data(groups = c(".model", "sex")) %>%
    flextable() %>%
    bold(i = ~ lb_pvalue < 0.05, 6) %>% #
    font(fontname = "sans", part = "all") %>% 
    add_header_lines("Check residuals with ljung box test, pvalue and stat")

plots <- pmap(list(p$models, p$sex, p$medicine), ~ ..1 %>% gg_tsresiduals(type = "innovation", lag_max = 24) + labs(title = paste("Residual plot for", model, ..2, ..3, sep = " ") ) )
  
```

```{r print_residual_check_table,  warning=FALSE}

for (i in 1:12) {
 plot.new()
 print(plots[[i]])
 }


ft

```

Plotting figure 1 and estimates of change with 95% confidence intervals

```{r seasonal_model_plot, eval = TRUE}

model <- c("seasonal_arma_log")


p <- filter_fun(model_ts, model)

# create flextable object for residual test table
tidy_estimates <- p %>%
  select(medicine, sex, tidy, test2) %>%
  unnest(c(tidy)) %>%
  select(medicine, sex, term, estimate, std.error, .model) %>%
  mutate(
    change = ifelse(grepl("log", .model) == TRUE & (term == "step" | term == "ramp" | term == "trend()"),
      # log model exponeniate if not just round:
      round(100 * (exp(estimate) - 1), digits = 2), round(estimate, digits = 2)
    ),
    lwr.ci = ifelse(grepl("log", .model) == TRUE & (term == "step" | term == "ramp" | term == "trend()"),
      round(100 * (exp(estimate - 1.96 * std.error) - 1), digits = 2), round(estimate - 1.96 * std.error, digits = 2)
    ),
    upr.ci = ifelse(grepl("log", .model) == TRUE & (term == "step" | term == "ramp" | term == "trend()"),
      round(100 * (exp(estimate + 1.96 * std.error) - 1), digits = 2), round(estimate + 1.96 * std.error, digits = 2)
    )
  ) %>%
  filter(
    term == "step" | term == "ramp" | term == "trend()"
  ) %>%
  mutate(
    term = ifelse(term == "trend()", "trend", term)
  ) %>%
  select(-std.error)


tidy_estimates$medicine <- factor(tidy_estimates$medicine, levels = c("Lisdexamfetamine", "Dexamfetamine", "Methylphenidate", "Nonstimulant"))
tidy_estimates$term <- factor(tidy_estimates$term, levels = c("step", "ramp", "trend"))


specs <- tidyr::build_wider_spec(tidy_estimates,
  names_from = term,
  values_from = c(change, lwr.ci, upr.ci),
  names_glue = "{term}_{.value}"
) %>%
  arrange(term, .value)


ft <- pivot_wider_spec(tidy_estimates %>% select(-estimate, -.model), specs) %>%
  arrange(sex, medicine) %>%
  as_grouped_data(groups = c("sex")) %>%
  unite("step_ci", c("step_lwr.ci", "step_upr.ci"), sep = " to ") %>%
  unite("ramp_ci", c("ramp_lwr.ci", "ramp_upr.ci"), sep = " to ") %>%
  unite("trend_ci", c("trend_lwr.ci", "trend_upr.ci"), sep = " to ") %>%
  flextable() %>%
  flextable::font(fontname = "sans", part = "all") %>%
  set_header_labels(sex = "Sex", medicine = "Medicine") %>%
  separate_header(split = "[_]", opts = c("center-hspan")) %>%
  autofit()%>% 
    add_header_lines("Time series estimates for each change effect, step, ramp, trend")


fit_specs <- p %>%
  select(medicine, sex, tidy, test2) %>%
  unnest(c(tidy)) %>%
  select(medicine, sex, term, estimate, .model) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(r = ramp, t = `trend()`, s = step)

df <- merge(ts, fit_specs, by = c("medicine", "sex")) %>% mutate(time = date - yearmonth(min(date)))

mean_data <- nest_ts %>%
  mutate(
    data = map(data, ~ .x %>%
      as_tibble() %>%
      select(date, rate, cnt, ramp, step) %>%
      mutate(time = date - yearmonth(min(date)))),


 # baseline for seasonal differences;
 
    data_mean = map_dbl(data, ~ .x %>%
      filter(time %in% c(0:11)) %>%
      mutate(mean = mean(log(rate))) %>%
      filter(time == 1) %>%
      select(mean) %>%
      pull())
  ) %>%
  select(data_mean, medicine, sex)


df <- merge(df, mean_data, by = c("medicine", "sex")) %>%
  rename(ss = data_mean) %>%
  mutate(
    step_value = s * step,
    ramp_value = r * ramp,
    # to account for seasonal differencing trend
    trend_value = t * (time - 6)
  ) %>%
  mutate(
    fit_un = rowSums(select(., .dots = all_of(c("step_value", "ramp_value", "trend_value", "ss"))), na.rm = TRUE),
    # exponentiating if needed i.e., model is log (all models?)
    fit = ifelse(grepl("log", .model) == TRUE, exp(fit_un), fit_un),
    class = factor(ifelse(medicine != "Lisdexamfetamine", "Other", "Lisdexamfetamine"))
  ) %>%
  mutate_at(1:2, factor) %>%
  ### creating a break in the trend line to improve readability of graph:
  # based on https://stackoverflow.com/questions/14821064/line-break-when-no-data-in-ggplot2

  mutate(fit = ifelse(date == yearmonth(change_date), NA_real_, fit))


plot <- df %>%
  filter(sex != "persons") %>%
  ggplot(aes(x = date, colour = medicine)) +
  geom_point(aes(x = date, y = rate), size = 1, na.rm = TRUE) +
  geom_line(aes(y = fit), alpha = .80, size = 1) +
  facet_grid(class ~ sex) +
  scale_x_yearmonth(date_labels = "%b %y", date_breaks = "1 year") +
  xlab("Date") +
  ylab("Monthly PBS dispensings per 1000 adults") +
  scale_y_continuous(limits = c(0, 2)) +
  geom_vline(xintercept = as.Date(change_date), linetype = 4, size = 1.1, colour = "black") +
  guides(colour = guide_legend(title = NULL)) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_line(colour = "grey90"),
    panel.grid.major.x = element_line(colour = "grey90"),
    panel.grid.minor.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "bottom",
    panel.border = element_rect(colour = "darkgrey", fill = NA), 
    axis.line = element_line(color = "black")
  )

ggsave(
  filename = paste(Sys.Date(), "Figure 1. A", model, "by medicine and sex.tiff"),
  plot = plot,
  device = NULL,
  scale = 1,
  # width = 4000,
  # height = 3000,
  units = "px",
  dpi = 300,
  limitsize = FALSE,
)


#print figure 1 
print(plot)

#print time series estimates

ft
 
```

Reporting mean rate of dispensing in 12 months prior to and following the intervention. 

```{r mean_before_after}

#based on data including April 2020 (removed from time series)

mean_data <- tss %>%
  inner_join(erpx, by = c("date", "sex")) %>%
  mutate(rate = (cnt * 10000) / value) %>%
  select(date, medicine, sex, rate) %>%
  filter_index( "2020-02-01"~"2022-01-31") %>% 
  mutate(year_time = ifelse(as.Date(date)<=as.Date("2021-01-31"), "mean_before", "mean_following")) %>% 
  as_tibble() %>%
  group_by(sex, medicine, year_time) %>%
    summarise(mean = mean(rate))

mean_data %>%
  pivot_wider(id_cols = c(sex, medicine), names_from = year_time, values_from = mean) %>%
  flextable() %>%
  add_header_lines("Mean monthly dispensings in year before and following intervention")

```

## References 

Key packages used are referenced here: 
Tidyverts: tsibble, fable, feasts
Wang, E, D Cook, and RJ Hyndman (2020). A new tidy data structure to support exploration and modeling of temporal data, Journal of Computational and Graphical Statistics, 29:3, 466-478, doi:10.1080/10618600.2019.1695624

O'Hara-Wild M, Hyndman R, Wang E (2022). _fable: Forecasting Models for Tidy Time Series_. R
  package version 0.3.2, <https://CRAN.R-project.org/package=fable>
  
O'Hara-Wild M, Hyndman R, Wang E (2022). _feasts: Feature Extraction and Statistics for Time
  Series_. R package version 0.3.0, <https://CRAN.R-project.org/package=feasts>
  
ggplot2:
Wickham H (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. ISBN 978-3-319-24277-4, https://ggplot2.tidyverse.org.

flextable:
Gohel D, Skintzos P (2023). flextable: Functions for Tabular Reporting. https://ardata-fr.github.io/flextable-book/, https://davidgohel.github.io/flextable/.

dplyr:
Wickham H, François R, Henry L, Müller K (2022). _dplyr: A Grammar of Data Manipulation_. R
  package version 1.0.10, <https://CRAN.R-project.org/package=dplyr>.
  
purrr:
 Wickham H, Henry L (2023). _purrr: Functional Programming Tools_. R package version 1.0.1,
  <https://CRAN.R-project.org/package=purrr>.
